---
title: "EDS 223 Homework Assignment 3"
author: "Rachel Swick"
date: last-modified
execute:
  warning: false
  message: false
format:
  html:
    toc: true
    code-fold: true
---

# Identifying the Impacts of Extreme Weather

### Learning Goals
In this assignment we will practice the following:
- load vector/raster data
- simple raster operations
- simple vector operations
- spatial joins

## 1. Load Packages and Import Data
```{r}
# Load packages
library(terra)
library(stars)
library(tidyverse)
library(tmap)
library(kableExtra) 
library(sf)
library(here)
library(testthat)
library(devtools)
library(remotes)
library(patchwork)
```

```{r}
# Load data
# 2021-02-07 night lights
night_lights_07_h08v05 <- rast("data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif")
night_lights_07_h08v06 <- rast("data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif")

# 2021-02-16 night lights
night_lights_16_h08v05 <- rast("data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif")
night_lights_16_h08v06 <- rast("data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif")

# Roads intersecting the Houston Metro Area
roads <- read_sf(here("data", "gis_osm_roads_free_1.gpkg"), 
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'")

# Houses in the Houston Metro Area
houses <- read_sf(here("data", "gis_osm_buildings_a_free_1.gpkg"),
                  query = "SELECT *
                          FROM gis_osm_buildings_a_free_1
                          WHERE (type IS NULL AND name IS NULL)
                          OR type in ('residential', 
                                      'apartments', 
                                      'house', 
                                      'static_caravan', 
                                      'detached')")

# Socioeconomic data
# Geometry layer
socio_geometry <- read_sf(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")
# Income layer
income <- read_sf(here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")
# Combine the geometry and income layers
income_geo <- left_join(socio_geometry, income, join_by(GEOID_Data == GEOID))
```


### 2. Ensure Matching Coordinate Reference Systems (CRS)

An accurate map can not be created without first ensuring all spatial objects are using the same CRS.

```{r}
# Create list of all spatial objects
spatial_objects <- list(night_lights_07_h08v05, 
                        night_lights_07_h08v06, 
                        night_lights_16_h08v05, 
                        night_lights_16_h08v06,
                        roads, 
                        houses, 
                        income_geo)

# Check if all spatial objects have the same CRS
# If not, determine whether it is a vector or raster object
# Update CRS based on object type
for (i in 2:length(spatial_objects)) {
  if (inherits(spatial_objects[[i]], "SpatRaster") && 
      !identical(crs(spatial_objects[[i]]), spatial_objects[[1]])) {
    spatial_objects[[i]] <- project(spatial_objects[[i]], crs(spatial_objects[[1]]))
    warning("The CRS has been transformed to match.")
  }
  else if (inherits(spatial_objects[[i]], c("sf", "tbl_df", "tbl", "data.frame")) &&
      st_crs(spatial_objects[[i]]) != st_crs(spatial_objects[[1]])) {
    spatial_objects[[i]] <- st_transform(spatial_objects[[i]], st_crs(spatial_objects[[1]]))
    warning("The CRS has been transformed to match.")
  }
  else {
    print("The CRS Match.")
  }
}

# Check that all spatial objects have been transformed
for (i in 2:length(spatial_objects)) {
  print(st_crs(spatial_objects[[i]]) == st_crs(spatial_objects[[1]]))
}
```

### 3. Create the Blackout Mask

In order to see which areas experienced blackout, we need to determine where there was a change in night light intensity pre and post storm. 

```{r}
# Combine the night light data for each date into one raster object
night_lights_07 <- merge(night_lights_07_h08v05, night_lights_07_h08v06)
night_lights_16 <- merge(night_lights_16_h08v05, night_lights_16_h08v06)

# Find the difference in night light intensity caused by the storm
night_light_diff <- night_lights_07 - night_lights_16
```

```{r}
# Assign NA to all locations that experienced a drop of less than 200 nW cm-2sr-1 change
night_light_diff[night_light_diff < 200 ] <- NA

# Vectorize the blackout mask
night_light_diff_vector <- as.polygons(night_light_diff) %>% 
  st_as_sf() %>% 
  st_make_valid()
```

```{r}
# Crop the blackout mask to the Houston area
# Create a bbox for the Houston area
houston <- st_bbox(c(xmin = -96.5,
                          ymin = 29,
                          xmax = -94.5,
                          ymax = 30.5)) %>% 
  st_as_sfc() %>% 
  st_set_crs(crs(night_light_diff_vector))

# Join the Houston bbox to the night light object
houston_night_lights_diff <- st_intersection(night_light_diff_vector, houston)
```

```{r}
# Change the crs of the Houston night light data set to EPSG:3083
houston_night_lights_diff <- st_transform(houston_night_lights_diff, crs = 3083)
```

### Map of the Houston Area Before the Storm
```{r}
# Light map before the storm
tm_shape(night_lights_07,
         bbox = houston) +
  tm_raster(palette = c("black", "#F2DD6E"), 
            breaks = c(0, 100, 200, 300, 500, 1000, 2000, 4000, 6000, 8000, Inf), 
            title = "Radiance (nW cm⁻² sr⁻¹)",
            legend.show = TRUE) +
  tm_layout(main.title = "Houston Metro Area Prior to 2017 Winter Storm",  
            main.title.size = 1.5,  
            main.title.position = c("center", "top"),  
            legend.outside = TRUE,  
            legend.outside.position = "right",
            attr.outside = TRUE) +
  tm_compass(position = c("RIGHT", "top")) +
  tm_scale_bar(position = c("LEFT", "top"))
```

### Map of the Houston Area After the Storm
```{r}
# Light map after the storm
tm_shape(night_lights_16,
         bbox = houston) +
  tm_raster(palette = c("black", "#F2DD6E"), 
            breaks = c(0, 100, 200, 300, 500, 1000, 2000, 4000, 6000, 8000, Inf), 
            title = "Radiance (nW cm⁻² sr⁻¹)", 
            legend.show = TRUE) +
  tm_layout(main.title = "Houston Metro Area After 2017 Winter Storm",  
            main.title.size = 1.5,  
            main.title.position = c("center", "top"),  
            legend.outside = TRUE,  
            legend.outside.position = "right",
            attr.outside = TRUE) +
  tm_compass(position = c("RIGHT", "top")) +
  tm_scale_bar(position = c("LEFT", "top"))
```

### Map of Areas in Houston that Experienced Power Outage
```{r}
tm_shape(houston_night_lights_diff) +
  tm_polygons(col = "black",
              border.col = "black") +
  tm_layout(main.title = "Houston Metro Areas Experiencing Blackout",  
            main.title.size = 1.5,  
            main.title.position = c("center", "top"),  
            legend.outside = TRUE,  
            legend.outside.position = "right",
            bg.color = "grey") +
  tm_compass(position = c("RIGHT", "bottom")) +
  tm_scale_bar(position = c("LEFT", "bottom")) +
  tm_add_legend(labels = "2017 Blackout Areas",
                col = "black")
```

### 4. Exclude highways from the cropped blackout mask

Since highways can have changes in light intensity unrelated to the winter storm, we need to exclude all areas within 200 meters of all highways in the Houston Metro area.
```{r}
# change the crs of roads to match the night lights data
roads <- st_transform(roads, crs = st_crs(houston_night_lights_diff))
# check the units of roads
st_crs(roads)$units
# identify areas within 200m of highways
roads_buffer <- st_buffer(roads, dist = 200) 

# Combine the buffered highways into a single object
roads_buffer_union <- st_union(roads_buffer)
```

```{r}
# Remove the highways from the night lights object
blackout <- st_difference(houston_night_lights_diff, roads_buffer_union)
```

### 5. Identify homes likely impacted by blackout

Identify homes that overlap with areas that experienced blackout.
```{r}
# Transform the CRS of the houses spatial object
houses <- st_transform(houses, crs = st_crs(houston_night_lights_diff))
# Homes affected by blackout
blackout_homes <- houses[blackout, ]
```

### Homes in Houston that Lost Power
```{r}
# Map of Houston area homes that lost power
tm_shape(roads) +
  tm_lines(col = "darkgrey",
           legend.col.show = TRUE) +
  tm_shape(blackout_homes) +
  tm_polygons(border.col = "#B87D4B")+
  tm_layout(main.title = "Houston Metro Area Homes that Lost Power",  
            main.title.size = 1.5,  
            main.title.position = c("center", "top"),
            legend.outside = TRUE,  
            legend.outside.position = "right") +
  tm_compass(position = c("RIGHT", "bottom")) +
  tm_scale_bar(position = c("LEFT", "bottom")) +
  tm_add_legend(labels = "Homes that Lost Power",
                col = "#B87D4B") +
  tm_add_legend(labels = "Roads",
                col = "darkgrey")
```

### Estimate of Houston Area Homes that Lost Power
```{r}
# Estimated number of homes that lost power
number_homes <- nrow(blackout_homes)

tribble(
  ~Description , ~Total,
  "Number of Homes", number_homes
) %>%
  kable()
```

### 6. Identify the Census Tracts Likely Impacted by Blackout

Identify census tracts that overlap with homes that experienced blackout.
```{r}
# Transform the CRS of the income_geo spatial object
income_geo <- st_transform(income_geo, crs = st_crs(houston_night_lights_diff))
# Census tracts affected by blackout
affected_census_tracts <- income_geo[blackout_homes, ]
```

### Census Tracts Impacted by Blackout
```{r}
# Map of census tracts affected by blackout
tm_shape(affected_census_tracts) +
  tm_polygons(col = "#210124",
              border.col = "grey",
              lwd = 0.05) +
  tm_layout(main.title = "Houston Area Census Tracts With Homes that Lost Power",  
            main.title.size = 1.5,  
            main.title.position = c("center", "top"),
            legend.outside = TRUE,  
            legend.outside.position = "right") +
  tm_compass(position = c("RIGHT", "bottom")) +
  tm_scale_bar(position = c("LEFT", "top")) +
  tm_add_legend(labels = "Census Tracts",
                col = "#210124",
                border.col = "grey") 
```

### Distributions of Median Household Incomes for Census Tracts that Did and Did Not Experience Blackout

Show the distributions of median household incomes for census tracts in which blackouts did and did not occur in order to determine if there is income inequity in where blackouts occurred.

```{r}
# Census tracts not affected by blackout
unaffected_census_tracts <- income_geo %>% 
  filter(!GEOID %in% affected_census_tracts$GEOID)
```

```{r}
# Distribution of median household income for census tracts that experienced blackout
affected <- ggplot(affected_census_tracts, aes(x = B19001e1)) +
  geom_histogram(fill = "#B3DEC1") +
  theme_classic() +
  labs(x = "Affected Census Tracts", y = "Number of Homes")

# Distribution of median household income for census tracts that did not experience blackout
unaffected <- ggplot(unaffected_census_tracts, aes(x = B19001e1)) +
  geom_histogram(fill = "#B3DEC1") +
  theme_classic() +
  labs(x = "Unaffected Census Tracts", y = "Number of Homes")

affected + unaffected + plot_annotation(
  title = "Median Household Income of Houston Area Census Tracts"
)
```


### Reflection

A significant number of homes in the Houston, TX metro area were impacted by blackout during and after the 2017 winter storms. The maps showing where blackouts occurred in the Houston area demonstrate that blackouts were not limited to any one area of the city and its surrounding neighborhoods. In addition, distributions of median household income in both affected and unaffected census tracts, show no clear differences. However, some census tracts are rather large and income inequalities might be more noticeable at a smaller scale. Additional analyzes may look at the racial makeup of census blocks to determine if there are other factors at play when it comes to who may be more suscpetible to blackouts in the future.

